---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
# IST 707 Assignment

### HW 1 - Association Rule Mining

#### 1) Data Loading
```{r}
library('arules')
data('AdultUCI')
```

#### 2) Data Exploration


##### Reviewing Structure and Summary of data
```{r}
str(AdultUCI)
summary(AdultUCI)
```
  
  
##### Viewing top entries in the data
```{r}
knitr::kable(head(AdultUCI))
```
  
##### Discretizing columns  
  
1. age
```{r}
q <- quantile(AdultUCI$age, c(0.4, 0.6))
AdultUCI$AgeBuckets <- replicate(length(AdultUCI$age), "Middle-aged")
AdultUCI$AgeBuckets[AdultUCI$age <= q[1]] <- "Young"
AdultUCI$AgeBuckets[AdultUCI$age > q[2]] <- "Old"
AdultUCI$AgeBuckets <- as.factor(AdultUCI$AgeBuckets)
str(AdultUCI$AgeBuckets)
```  
2. fnlwgt
```{r}
q <- quantile(AdultUCI$fnlwgt, c(0.4, 0.6))
AdultUCI$fnlwgtBuckets <- replicate(length(AdultUCI$fnlwgt), "Average")
AdultUCI$fnlwgtBuckets[AdultUCI$fnlwgt <= q[1]] <- "Low"
AdultUCI$fnlwgtBuckets[AdultUCI$fnlwgt > q[2]] <- "High"
AdultUCI$fnlwgtBuckets <- as.factor(AdultUCI$fnlwgtBuckets)
str(AdultUCI$fnlwgtBuckets)
```  
3. capital-gain
```{r}
q <- quantile(AdultUCI$`capital-gain`, c(0.4, 0.6))
AdultUCI$CapitalGainBuckets <- replicate(length(AdultUCI$`capital-gain`), "Average")
AdultUCI$CapitalGainBuckets[AdultUCI$`capital-gain` <= q[1]] <- "Low"
AdultUCI$CapitalGainBuckets[AdultUCI$`capital-gain` > q[2]] <- "High"
AdultUCI$CapitalGainBuckets <- as.factor(AdultUCI$CapitalGainBuckets)
str(AdultUCI$CapitalGainBuckets)
```  
4. capital-loss
```{r}
q <- quantile(AdultUCI$`capital-loss`, c(0.4, 0.6))
AdultUCI$CapitalLossBuckets <- replicate(length(AdultUCI$`capital-loss`), "Average")
AdultUCI$CapitalLossBuckets[AdultUCI$`capital-loss` <= q[1]] <- "Low"
AdultUCI$CapitalLossBuckets[AdultUCI$`capital-loss` > q[2]] <- "High"
AdultUCI$CapitalLossBuckets <- as.factor(AdultUCI$CapitalLossBuckets)
str(AdultUCI$CapitalLossBuckets)
```  
5. hours-per-week
```{r}
q <- quantile(AdultUCI$`hours-per-week`, c(0.4, 0.6))
AdultUCI$HoursPerWeekBuckets <- replicate(length(AdultUCI$`hours-per-week`), "Average")
AdultUCI$HoursPerWeekBuckets[AdultUCI$`hours-per-week` <= q[1]] <- "Low"
AdultUCI$HoursPerWeekBuckets[AdultUCI$`hours-per-week` > q[2]] <- "High"
AdultUCI$HoursPerWeekBuckets <- as.factor(AdultUCI$HoursPerWeekBuckets)
str(AdultUCI$HoursPerWeekBuckets)
```

##### Checking for duplicated data
```{r}
head(AdultUCI[duplicated(AdultUCI),],2)
```
##### Removing duplicate data
```{r}
AdultUCI_WithoutDuplicates <- AdultUCI[!duplicated(AdultUCI),]
nrow(AdultUCI) #with duplicates
nrow(AdultUCI_WithoutDuplicates) #without duplicates
```

##### Checking for NA values
```{r}

colnames(AdultUCI_WithoutDuplicates)[colSums(is.na(AdultUCI_WithoutDuplicates)) > 0]

```
  
##### Replacing NAs
```{r}
AdultUCI_WithoutDuplicates[is.na(AdultUCI_WithoutDuplicates$workclass),]$workclass <- 'no workclass'
AdultUCI_WithoutDuplicates[is.na(AdultUCI_WithoutDuplicates$occupation), ]$occupation <- 'no occupation'
AdultUCI_WithoutDuplicates[is.na(AdultUCI_WithoutDuplicates$`native-country`), ]$`native-country` <- 'United-States'
AdultUCI_WithoutDuplicates[is.na(AdultUCI_WithoutDuplicates$income), ]$income <- 'no income'
```


##### Setting final dataset
```{r}
AdultUCI_WithoutDuplicates$`education-num` <- as.factor(AdultUCI_WithoutDuplicates$`education-num`)
AdultUCIFinal <- AdultUCI_WithoutDuplicates[,c(2,4,5,6,7,8,9,10,14,15,16,17,18,19,20)]
```
#### 3) Running apriori with default settings
```{r}
AdultUCI_Rules_1 <- apriori(AdultUCIFinal)
AdultUCI_Rules_1 <- sort (AdultUCI_Rules_1, by="confidence",decreasing=TRUE)
inspect(head(AdultUCI_Rules_1,5) )
```
#### 4) Running apriori with support = 0.1 and confidence = 0.5
```{r}
AdultUCI_Rules_2 <- apriori(AdultUCIFinal, parameter = list(supp = 0.1, conf = 0.5))
AdultUCI_Rules_2 <- sort (AdultUCI_Rules_2, by="confidence",decreasing=TRUE)
inspect(head(AdultUCI_Rules_2,5) )
```
#### 5) Running apriori with support = 0.5 and confidence = 0.5
```{r}
AdultUCI_Rules_3 <- apriori(AdultUCIFinal, parameter = list(supp = 0.5, conf = 0.5))
AdultUCI_Rules_3 <- sort (AdultUCI_Rules_3, by="confidence",decreasing=TRUE)
inspect(head(AdultUCI_Rules_3,5) )
```
#### 6) Running apriori to predict income
```{r}
AdultUCI_Rules_4 <- apriori(AdultUCIFinal, parameter=list (supp=0.1,conf = 0.5, maxtime = 10), appearance = list (default="lhs",rhs=c("income=small","income=large")))
AdultUCI_Rules_4 <- sort (AdultUCI_Rules_4, by="confidence",decreasing=TRUE)
inspect(head(AdultUCI_Rules_4,5))
```
##### Evaluating rules :  The first few rules with highest confidence state that when marital-status is 'never married', age is 'young', capital gain and loss is 'low', and hours per week is 'low', then the income is 'low'. This rule makes sense because a young person working for less hours will recieve a low income.

#### 7) Shiny app for Apriori
#####https://achandir.shinyapps.io/Assignment1/